---
title: "Defining parameters to identify isotopologue peaks using compounds from HMDB"
author: "Andrea Vicini, Vinicius Verri Hernandez, Johannes Rainer"
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    toc_depth: 3
    fig_width: 5
---

```{r style, echo = FALSE, results = 'asis', message = FALSE}
library(BiocStyle)
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

**Last modified:** `r file.info("hmdb_subst_neutral.Rmd")$mtime`<br />
**Compiled**: `r date()`

$\require{mhchem}$

# Introduction

In this document we define parameters to identify and group potential
isotopologue peaks based on compound definitions from the Human Metabolome
Database [HMDB](https://hmdb.ca). We use the chemical formulas of the compounds
as they are provided by HMDB hence estimating all parameters for the *neutral*
compounds (not specific ions).

```{r include_approach, echo = FALSE, results = "asis"}
cat(readLines("./approach.md"), sep = "\n")
```


# Parameter estimation

Before proceeding to the actual estimation of the parameters we evaluate element
counts for all compounds listed in HMDB.

```{r libraries, warning = FALSE}
library("CompoundDb")
library("MetaboCoreUtils")
library("pander")
library("enviPat")
source("functions.R")
```

We load the HMDB database and extract the chemical formula and exact mass of all
compounds.

```{r}
cdb <- CompDb("data/CompDb.Hsapiens.HMDB.4.0.sqlite")
cmps <- compounds(cdb, columns = c("compound_id", "name",
                                   "formula", "exactmass"))
```

For each compound in the human metabolom database (HMDB) with available mass 
we compute how many atoms of C, H, N, O, P and S as well as Cl are present in it
(the latter is important because in human serum samples it is very likely that
Cl adducts are generated by electro spray ionization).
We collect the counts in a data frame with a column for each element and a row 
for each compound in HMDB.

```{r}
cmps <- cmps[-which(is.na(cmps$exactmass)), ]
```

```{r}
tmp <- lapply(cmps$formula, function(frml) res <- countElements(frml))
elements_names <- unique(names(unlist(tmp)))
counts <- lapply(tmp, function(x) {
  row <- numeric(length(elements_names))
  names(row) <- elements_names
  row[names(x)] <- x
  row
  })
counts_df <- data.frame(do.call(rbind, counts))
rownames(counts_df) <- cmps$compound_id
```

```{r}
CHNOPSCl <- c("C", "H", "N", "O", "P", "S", "Cl")
counts_CHNOPSCl <- data.frame(counts_df[ , CHNOPSCl])
```

```{r}
onlyCHNOPSCl <- rowSums(counts_df[, setdiff(elements_names, CHNOPSCl)]) == 0

# With the following lines we can exclude compounds that have some atoms of
# elements different from CHNOPSCl.
cmps <- cmps[onlyCHNOPSCl, ]
counts_df <- counts_df[onlyCHNOPSCl, ]

```

From the `r nrow(counts_df)` compounds in HMDB
`r sum(onlyCHNOPSCl != 1)` contain elements other than
`r paste(CHNOPSCl, collapse = ",")`.

We compute the mean and standard deviation for the number of each element in the
molecules of HMDB. For each element we also report the percentage of compounds
which have at least one atom of the given element.

```{r, echo = FALSE, results = "asis"}
m_sd <- rbind(colMeans(counts_CHNOPSCl), 
              apply(counts_CHNOPSCl, 2, sd),
              100 * colSums(counts_CHNOPSCl > 0) / nrow(counts_CHNOPSCl))
rownames(m_sd) = c("mean", "standard deviation", "% compounds with count > 0")
pandoc.table(m_sd, style = "rmarkdown")
```

We plot, for each of the considered elements, the distribution of the number of
atoms of the given element in the compounds of HMDB.

```{r distribution-counts, fig.cap = "Distribution of the number of atoms of CHNOPSCl elements in compounds of HMDB.", fig.width = 7, fig.height = 14, echo = FALSE}
par(mfrow = c(7, 1), mar = c(3, 4, 1, 0.5))
for (el in CHNOPSCl) {
  hist(counts_CHNOPSCl[, el], xlab = "", main = el, breaks = 256)
  abline(v = quantile(counts_CHNOPSCl[, el], c(0.025, 0.975)))
}
```

The most frequent elements in HMDB compounds are H, C and O, but while the
number of C and H atoms varies considerably and goes up to over 100, the maximal
number of O atoms if below 25. Also, the numbers of atoms for all other elements
is only very low.



## Identification of the most frequent isotopic substitutions {#isotopologues}

We next aim at identifying, for all compounds present in HMDB, the most
frequently observed isotopic substitutions considering all possible
isotopologues for each molecule.

To this end we calculate the isotope pattern for each compound in HMDB using
`enviPat` and identify isotopologues which would result in an intensity
higher than 10^(-4). For each of these peaks we define its *substitution id* which
consists of the heavier isotopes characterizing the associated isotopologue,
e.g. `"[13]C4[15]N2"` and determine the frequency of each substitution in HMDB.

```{r, echo = FALSE}
dr <- paste0("data/RData/")
dir.create(dr, showWarnings = FALSE, recursive = TRUE)
```

```{r}
data(isotopes)
isos <- isotopes$isotope[-match(unique(isotopes$element), isotopes$element)]
```


```{r, eval = !file.exists(paste0(dr, "subst_frequencies_neutral.RData"))}
## Ensure that element order is correct
chemforms <- sapply(cmps$formula, standardizeFormula)
threshold <- 0.01 * 10^(-2)
iso_ps <- isopattern(isotopes, chemforms , threshold = threshold, rel_to = 2)

## Define the isotopologue name for each substitution - ensuring the order of
## isotopes to be the same across compounds
iso_ps <- lapply(iso_ps, function(iso_p) {
    idx <- match(colnames(iso_p), isos)
    tmp <- iso_p[, !is.na(idx), drop = FALSE]
    mono_iso <- rowSums(tmp) == 0
    tmp <- tmp[!mono_iso, order(idx[!is.na(idx)]), drop = FALSE]
    cn <- colnames(tmp)
    elname <- gsub('[0-9]', '', cn)
    isonumb <- gsub('[^0-9]', '', cn)
    nms <- rep("", nrow(iso_p))
    nms[!mono_iso] <- apply(tmp, 1, function(row)
        paste0("[", isonumb[row >0],  "]", elname[row > 0], row[row > 0],
               collapse = ""))
    nms[mono_iso] <- "mono"
    rownames(iso_p) <- nms
    iso_p
})

subst_per_cmpd <- lapply(iso_ps, function(z) rownames(z)[rownames(z) != "mono"])
subst_frequency <- table(unlist(subst_per_cmpd, use.names = FALSE))
save(subst_frequency, subst_per_cmpd,
     iso_ps, chemforms, file = paste0(dr, "subst_frequencies_neutral.RData"))
```


```{r, echo = FALSE, eval = file.exists(paste0(dr, "subst_frequencies_neutral.RData"))}
load(paste0(dr, "subst_frequencies_neutral.RData"))
```

We thus defined `r length(subst_frequency)` substitutions that in at least a 
compound in HMBD would yield a probability higher than 10^(-4).

For each given substitution we compute the mean intensity it has in the
compounds of HMBD where it was found to result in a peak higher then the set
threshold.

```{r}
intens <- unlist(lapply(iso_ps, function(iso_p) {
    iso_p[rownames(iso_p) != "mono", "abundance"]
}), use.names = FALSE)
m_intens_subs <- tapply(intens, unlist(subst_per_cmpd, use.names = FALSE), mean)
```

and plot it against the frequency of the substitution in HMBD. Most of the 
substitution present a small mean intensity. Most of the substitution with mean 
intensity higher than 5 are those related to Cl (these are not very frequent) 
and C.

```{r, fig.height = 5, fig.width = 9, fig.cap = "Substitutions frequencies and mean intensities."}
library(ggplot2)
library(plotly)
p <- data.frame(subst_name = names(m_intens_subs),
                subst_frequency = as.numeric(subst_frequency),
                mean_intensity = m_intens_subs) %>%
  ggplot(aes(subst_frequency, mean_intensity, label = subst_name)) + geom_point()
ggplotly(p + labs(x = "frequency", y = "mean intensity"))
```

We next identify the *most frequently* observed substitutions in compounds of
HMDB. A first approach could be simply to select the substitutions found in 50% 
of compounds to yield an intensity larger than the set threshold. 
Because of the high frequency of C, H and O elements in compounds of HMDB, this 
selection is heavily biased by isotopes of these elements.

To correct this bias we divide the number of times a substitution was found
*significant* (i.e. with an intensity above the threshold) with the number of
compounds in the database with a chemical formula that contains at least all
elements (and their counts) of the isotopic substitution. In other words these
ratios represent the likelihood to find a substitution significant among the
subset of HMBD compounds that have it as possible substitution (i.e. for which
the number of elements from the isotopic substitution is equal or larger then
the total sum of heavier isotopes for these elements in the substitution).

```{r}
nvar <- sapply(names(subst_frequency), function(subst) {
  els <- count_elements(subst)
  var <- rep(TRUE, nrow(counts_df))
  for (el in names(els))
    var <- var & (counts_df[, el] >= els[el])
  sum(var)
})
```

```{r, fig.height = 5, fig.width = 17, fig.cap = "Substitutions with corrected proportion > 0.5"}
subst_proportion_var <- sort(subst_frequency/nvar, decreasing = TRUE)
par(mar = c(10, 4.1, 4.1, 2.1))
barplot(subst_proportion_var[subst_proportion_var > 0.5], las = 2)
```

In the plot above we can see the substitutions that produce a significant 
peak for at least a half of the compounds in which they are possible. 
(the height of the bar specifies the exact proportion). In this case we observe 
also substitutions presenting N, S or Cl atoms. 
For example, all molecules (the proportion in the above plot is 
1) of HMBD that have at least 1 atom of C and 1 atom of Cl are found to have 
significant peaks corresponding to [13]C1[37]Cl1 despite the fact that the 
number of such compounds in HMBD is very low.

In the following plot we show also the mean intensity of the substitutions.  We
notice that substitutions of Cl which have high mean intensity are selected (but
would be excluded considering a simple proportion).

```{r, fig.height = 5, fig.width = 9, fig.cap = "Corrected proportions and mean intensities for the substitutions. The substitutions on the right of the red line correspond to the columns in the previous barplot"}
library(ggplot2)
library(plotly)
par(mfrow= c(1,2))
p <- data.frame(subst_name = names(m_intens_subs),
                proportion_var = as.numeric(subst_frequency/nvar),
                mean_intensity = m_intens_subs) %>%
ggplot(aes(proportion_var, mean_intensity, label = subst_name)) + geom_point()
ggplotly(p + geom_vline(xintercept = 0.5, color = "red", size= .1) +
           labs(x = "proportion (corrected)", y = "mean intensity"))
```

```{r}
selected_substs <- names(subst_proportion_var[subst_proportion_var > 0.5])
```

## Defining mass difference and intensity (probability) ratios

We next compute for all of the isotopologues (associated with the selected 
substitutions) their mass difference $md$ and subsequently also the observed 
intensity (probability) ratios between monoisotopic peak and isotopologue peak.

```{r, eval = !file.exists(paste0(dr, "iso_data_sub_neutral.RData"))}
## Calculate mass differences and intensity (probability) ratios for all
## isotopologues; putting all into a data.frame for easier data processing
iso_data <- lapply(iso_ps, function(z) {
    mono <- which(rownames(z) == "mono")
    if (length(mono) == 1L) {
        res <- data.frame(isotopologue = rownames(z),
                          md = z[, "m/z"] - z[mono, "m/z"],
                          R = z[, "abundance"] / z[mono, "abundance"],
                          z[, c("m/z", "abundance"), drop = FALSE],
                          monomass = z[mono, "m/z"],
                          check.names = FALSE)
        res[-mono, , drop = FALSE]
    } else {
        data.frame(isotopologue = rownames(z),
                   md = NA_real_,
                   R = NA_real_,
                   z[, c("m/z", "abundance"), drop = FALSE],
                   monomass = NA_real_,
                   check.names = FALSE)
    }
})
cnts <- vapply(iso_data, nrow, integer(1))
iso_data <- do.call(rbind, iso_data)
rownames(iso_data) <- NULL
iso_data$formula <- rep(cmps$formula, cnts)
iso_data$hmdb_id <- rep(cmps$compound_id, cnts)

## subset to selected isotopologues
iso_data_sub <- iso_data[iso_data$isotopologue %in% selected_substs, ]
save(iso_data_sub, file = paste0(dr, "iso_data_sub_neutral.RData"))
```

```{r, eval = file.exists(paste0(dr, "iso_data_sub_neutral.RData"))}
load(paste0(dr, "iso_data_sub_neutral.RData"))
```


## Defining mass-dependend lower and upper bounds for intensity ratios

Based on the available intensity ratios calculated on all compounds from HMDB we
can define upper and lower observed ratios that could be used to determine
whether the observed intensity ratio of a candidate isotopologue peak would fit
the expected ratio. Since the intensity ratio $R$ of the isotopic substitutions
is dependent on the number of elements within the compound and hence also on the
absolute mass of the compound, we define mass-dependent lower and upper
intensity ratios.
We thus construct for each isotopic substitution segmented upper and lower bound
lines based on the mass-to-$R$ relationships of all HMDB compounds
for which this substitution was found. To determine the points characterizing
the segmented upper [lower] bound we use an iterative approach whose fundamental
idea is to select the next point $P_{i+1}$ as the point that maximizes[minimizes] 
the slope of the line through the current point $P_{i}$ and $P_{i+1}$ over the
points with mass in a certain interval having left endpoint equal to the mass of 
$P_{i}$ (we use the the point (0, 0) as the starting point of the algorithm).
Using this approach we compute upper and lower bound broken lines
and we collect slope and intercepts of each segment in a substitution definition
matrix (along with the mass difference associated to the substitution). For
the upper bound computation the modified approach has been used with a x axis
subdivision presenting shorter intervals the closer to the minimum of the mass
range. For each substitution multiple rows are present and each one of them is
associated to a different mass interval whose endpoints are the endpoints
x-values of a segment in the broken lines.

```{r}
hmdb_subst <- lapply(selected_substs, function(sub){
  idx <- which(iso_data_sub$isotopologue == sub)
  points <- cbind(iso_data_sub$monomass[idx], iso_data_sub$R[idx])
  xsubd <- seq(0,1, length.out = 8)^2 * diff(range(points[, 1])) +
    min(points[, 1])
  bpl <- bound_points(points, "lower")
  bpu <- bound_points2(points, "upper", xsubd = xsubd)
  res <- get_bounds(bpl, bpu)
  nr <- nrow(res)
  data.frame(name = rep(sub, nr),
             md = rep(iso_data_sub$md[idx][1], nr),
             minmass = rep(min(points[, 1]), nr),
             maxmass = rep(max(points[, 1]), nr),
             res)
})
hmdb_subst <- do.call(rbind, hmdb_subst)
hmdb_subst <- hmdb_subst[order(hmdb_subst[, "md"]), ]
```

In the following plot are reported the obtained bounds for R vs mass.

```{r, echo = FALSE, fig.width = 10, fig.height = 8}
f <- rep(seq(ceiling(length(selected_substs)/9)),
         each = 9)[seq_along(selected_substs)]
plts <- split(sort(selected_substs), f)

tmp <- lapply(plts, function(z) {
  par(mfrow = c(3, 3), mar = c(4.2, 4.2, 2, 0.5))
  for (iso in z) {
    idx <- which(iso_data_sub$isotopologue == iso)
    points <- cbind(iso_data_sub$monomass[idx], iso_data_sub$R[idx])
    plot(points, main = iso, xlab = "monoisotopic mass",
         ylab = "R", pch = 16, col = "#00000040", cex = 0.75)
    idx_sub <- which(hmdb_subst[, "name"] == iso)
    x <- hmdb_subst[idx_sub, "rightend"]
    LBint <-  hmdb_subst[idx_sub, "LBint"]
    LBslope <-  hmdb_subst[idx_sub, "LBslope"]
    points(cbind(c(0, x), c(0, LBslope * x + LBint)), type = "l", col = "#377EB880")
    UBint <-  hmdb_subst[idx_sub, "UBint"]
    UBslope <-  hmdb_subst[idx_sub, "UBslope"]
    points(cbind(c(0, x), c(0, UBslope * x + UBint)), type = "l", col = "#4DAF4A80")
    abline(v = c(0, x), col = "grey", lwd = 0.5)
  }
})
```

Segmentation along the mass (m/z) axis allows also to describe the upper
and lower limits for non-linear relationships between intensity ratio $R$ and
mass.

We repeat the same plots on a portion of the mass range.

```{r, echo = FALSE, fig.width = 10, fig.height = 8}
f <- rep(seq(ceiling(length(selected_substs)/9)),
         each = 9)[seq_along(selected_substs)]
plts <- split(sort(selected_substs), f)

tmp <- lapply(plts, function(z) {
  par(mfrow = c(3, 3), mar = c(4.2, 4.2, 2, 0.5))
  for (iso in z) {
    idx <- which(iso_data_sub$isotopologue == iso)
    points <- cbind(iso_data_sub$monomass[idx], iso_data_sub$R[idx])
    #s <- which(points[, 1] <= 800)
    s <- which(points[, 1] <= min(points[, 1]) + 700)
    plot(points[s, ], main = iso, xlab = "monoisotopic mass",
         ylab = "R", pch = 16, col = "#00000040", cex = 0.75)
    idx_sub <- which(hmdb_subst[, "name"] == iso)
    x <- hmdb_subst[idx_sub, "rightend"]
    LBint <-  hmdb_subst[idx_sub, "LBint"]
    LBslope <-  hmdb_subst[idx_sub, "LBslope"]
    points(cbind(c(0, x), c(0, LBslope * x + LBint)), type = "l", col = "#377EB880")
    UBint <-  hmdb_subst[idx_sub, "UBint"]
    UBslope <-  hmdb_subst[idx_sub, "UBslope"]
    points(cbind(c(0, x), c(0, UBslope * x + UBint)), type = "l", col = "#4DAF4A80")
    abline(v = c(0, x), col = "grey", lwd = 0.5)
  }
})
```

```{r}
dr_txt <- paste0("data/txt/")
dir.create(dr_txt, showWarnings = FALSE, recursive = TRUE)
write.table(hmdb_subst, file = paste0(dr_txt, "hmdb_subst_neutral.txt"),
            sep = "\t", row.names = TRUE)
```

The following table contains summary information about the substitutions.

```{r, echo = FALSE, results = "asis"}
f <- factor(iso_data_sub$isotopologue)
subst_info <- data.frame(
  name = levels(f),
  minmass = vapply(split(iso_data_sub$monomass, f), min, numeric(1)),
  maxmass = vapply(split(iso_data_sub$monomass, f), max, numeric(1)),
  md = iso_data_sub$md[match(levels(f), iso_data_sub$isotopologue)],
  proportion = as.numeric(
    (subst_frequency / nrow(cmps))[match(levels(f), names(subst_frequency))]),
  proportion_var = as.numeric(
    subst_proportion_var[match(levels(f), names(subst_proportion_var))])
)
subst_info <- subst_info[order(subst_info$md), ]
rownames(subst_info) <- NULL
pandoc.table(subst_info, style = "rmarkdown", split.tables = Inf)
```


# Session information

The R version and packages used in this analysis are listed below.

```{r sessioninfo}
sessionInfo()
```

Information on the HMDB data used:

```{r}
metadata(cdb)
```
