---
title: "Defining parameters to identify isotopologue peaks using [M+H], [M+k], [M+Na] from HMDB"
author: "Andrea Vicini, Vinicius Verri Hernandez, Johannes Rainer"
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    toc_depth: 3
    fig_width: 5
---

```{r style, echo = FALSE, results = 'asis', message = FALSE}
library(BiocStyle)
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
```

**Last modified:** `r file.info("hmdb_subst_positive.Rmd")$mtime`<br />
**Compiled**: `r date()`

$\require{mhchem}$

# Introduction

In this document we evaluate the substitutions parameters that can be obtained 
from the the adducts [M+H], [M+k], [M+Na] of the human metabolom database 
(HMDB) compounds.

```{r libraries, warning = FALSE}
library("CompoundDb")
library("MetaboCoreUtils")
library("pander")
library("enviPat")
source("functions.R")
```

We load the HMDB database and extract the chemical formula and exact mass of all
compounds.

```{r}
cdb <- CompDb("data/CompDb.Hsapiens.HMDB.4.0.sqlite")
cmps <- compounds(cdb, columns = c("compound_id", "name",
                                   "formula", "exactmass"))
```

For each compound in the HMDB with available mass we compute the counts of 
the atoms that are present in it and collect them in a data frame with a 
column for each element and a row for each compound in HMDB.

```{r}
cmps <- cmps[-which(is.na(cmps$exactmass)), ]
```

```{r}
tmp <- lapply(cmps$formula, function(frml) res <- countElements(frml))
elements_names <- unique(names(unlist(tmp)))
counts <- lapply(tmp, function(x) {
  row <- numeric(length(elements_names))
  names(row) <- elements_names
  row[names(x)] <- x
  row
  })
counts_df <- data.frame(do.call(rbind, counts))
rownames(counts_df) <- cmps$compound_id
```

```{r}
CHNOPSCl <- c("C", "H", "N", "O", "P", "S", "Cl")
onlyCHNOPSCl <- rowSums(counts_df[, setdiff(elements_names, CHNOPSCl)]) == 0

# With the following lines we can exclude compounds that have some atoms of
# elements different from CHNOPSCl.
cmps <- cmps[onlyCHNOPSCl, ]
counts_df <- counts_df[onlyCHNOPSCl, ]
```

We derive the counts of the adducts [M+Cl] and from them the their formulas.

```{r}
counts_df_H <- counts_df_K <- counts_df_Na <- counts_df
counts_df_H$H <- counts_df$H + 1
counts_df_K$K <- counts_df$K + 1
counts_df_Na$Na <- counts_df$Na + 1
counts_df <- rbind(counts_df_H, counts_df_K, counts_df_Na)
adducts_frmls <- apply(counts_df, 1, function(elements) pasteElements(elements))
```

### Identification of the most frequent isotopic substitutions {#isotopologues}

We next aim at identifying, for all compounds present in HMDB, the most 
frequently observed isotopic substitutions considering all possible 
isotopologues for each molecule.

To this end we calculate the isotope pattern for each compound in HMDB using
`enviPat` and identify isotopologues which would result in an intensity
higher than 10^(-4). For each of these peaks we define its *substitution id* which
consists of the heavier isotopes characterizing the associated isotopologue,
e.g. `"[13]C4[15]N2"` and determine the frequency of each substitution in HMDB.

```{r, echo = FALSE}
dr <- paste0("data/RData/")
dir.create(dr, showWarnings = FALSE, recursive = TRUE)
```

```{r}
data(isotopes)
isos <- isotopes$isotope[-match(unique(isotopes$element), isotopes$element)]
```

```{r, eval = !file.exists(paste0(dr, "subst_frequencies_positive.RData"))}
## Ensure that element order is correct
chemforms <- sapply(adducts_frmls, standardizeFormula)
threshold <- 0.01 * 10^(-2)
iso_ps <- isopattern(isotopes, chemforms , threshold = threshold, rel_to = 2) 

## Define the isotopologue name for each substitution - ensuring the order of
## isotopes to be the same across compounds
iso_ps <- lapply(iso_ps, function(iso_p) {
    idx <- match(colnames(iso_p), isos)
    tmp <- iso_p[, !is.na(idx), drop = FALSE]
    mono_iso <- rowSums(tmp) == 0
    tmp <- tmp[!mono_iso, order(idx[!is.na(idx)]), drop = FALSE]
    cn <- colnames(tmp)
    elname <- gsub('[0-9]', '', cn)
    isonumb <- gsub('[^0-9]', '', cn)
    nms <- rep("", nrow(iso_p))
    nms[!mono_iso] <- apply(tmp, 1, function(row)
        paste0("[", isonumb[row > 0],  "]", elname[row > 0], row[row > 0], 
               collapse = ""))
    nms[mono_iso] <- "mono"
    rownames(iso_p) <- nms
    iso_p
})

subst_per_cmpd <- lapply(iso_ps, function(z) rownames(z)[rownames(z) != "mono"])
subst_frequency <- table(unlist(subst_per_cmpd, use.names = FALSE))
save(subst_frequency, subst_per_cmpd, 
     iso_ps, chemforms, file = paste0(dr, "subst_frequencies_positive.RData"))
```


```{r, echo = FALSE, eval = file.exists(paste0(dr, "subst_frequencies_positive.RData"))}
load(paste0(dr, "subst_frequencies_positive.RData"))
```

We thus defined `r length(subst_frequency)` substitutions that in at least a 
compound in HMBD would yield a probability higher than 10^(-4).

For each given substitution we compute the mean intensity it has in the
compounds of HMBD where it was found to result in a peak higher then the set
threshold.

```{r}
intens <- unlist(lapply(iso_ps, function(iso_p) {
    iso_p[rownames(iso_p) != "mono", "abundance"]
}), use.names = FALSE)
m_intens_subs <- tapply(intens, unlist(subst_per_cmpd, use.names = FALSE), mean)
```

and plot it against the frequency of the substitution in HMBD. Most of the 
substitution present a small mean intensity. Most of the substitution with mean 
intensity higher than 5 are those related to Cl (these are not very frequent) 
and C.

```{r, fig.height = 5, fig.width = 9, fig.cap = "Substitutions frequencies and mean intensities."}
library(ggplot2)
library(plotly)
p <- data.frame(subst_name = names(m_intens_subs),
                subst_frequency = as.numeric(subst_frequency), 
                mean_intensity = m_intens_subs) %>%
  ggplot(aes(subst_frequency, mean_intensity, label = subst_name)) + geom_point()
ggplotly(p + labs(x = "frequency", y = "mean intensity"))
```

We next identify the *most frequently* observed substitutions in compounds of
HMDB. A first approach could be simply to select the substitutions found in 50% 
of compounds to yield an intensity larger than the set threshold. 
Because of the high frequency of C, H and O elements in compounds of HMDB, this 
selection is heavily biased by isotopes of these elements.

To correct this bias we divide the counts of the times a substitution is found 
to be significant in the HMBD compounds for the times that this substitution 
can be observed across the HMBD compounds. In other words these ratios represent 
the likelihood to find a substitution significant among the subset of HMBD 
compounds that have it as possible substitution (this happens when, for each 
element $X$ in the compound, the number of atoms of $X$ is >= than the sum of 
the numbers ${n_X}_2$, ..., ${n_X}_q$ of heavier isotopes associated to the 
substitution).

```{r}
nvar <- sapply(names(subst_frequency), function(subst) {
  els <- count_elements(subst)
  var <- rep(TRUE, nrow(counts_df))
  for (el in names(els)) 
    var <- var & (counts_df[, el] >= els[el])
  sum(var)
})
```

```{r, fig.height = 5, fig.width = 17, fig.cap = "Substitutions with corrected proportion > 0.5"}
subst_proportion_var <- sort(subst_frequency/nvar, decreasing = TRUE)
par(mar = c(10, 4.1, 4.1, 2.1))
barplot(subst_proportion_var[subst_proportion_var > 0.5], las = 2)
```

In the plot above we can see the substitutions that produce a significant 
peak for at least a half of the compounds in which they are possible. 
(the height of the bar specifies the exact proportion). In this case we observe 
also substitutions presenting N, S or Cl atoms. 
For example, all molecules (the proportion in the above plot is 
1) of HMBD that have at least 1 atom of C, S and Cl are found to have 
significant peaks corresponding to [13]C1[33]S1[37]Cl1 despite the fact that the 
number of such compounds in HMBD is very low.

In the following plot we show also the mean intensity of the substitutions. 
We notice that substitutions of Cl which have high 
mean intensity are selected (but would be excluded considering a simple 
proportion).

```{r, fig.height = 5, fig.width = 9, fig.cap = "Corrected proportions and mean intensities for the substitutions. The substitutions on the right of the red line correspond to the columns in the previous barplot"}
library(ggplot2)
library(plotly)
par(mfrow= c(1,2))
p <- data.frame(subst_name = names(m_intens_subs),
                proportion_var = as.numeric(subst_frequency/nvar), 
                mean_intensity = m_intens_subs) %>%
ggplot(aes(proportion_var, mean_intensity, label = subst_name)) + geom_point()
ggplotly(p + geom_vline(xintercept = 0.5, color = "red", size= .1) + 
           labs(x = "proportion (corrected)", y = "mean intensity"))
```

```{r}
selected_substs <- names(subst_proportion_var[subst_proportion_var > 0.5])
```

### Defining mass difference and intensity (probability) ratios

We next compute for all of the isotopologues (associated with the selected 
substitutions) their mass difference $md$ and subsequently also the observed 
intensity (probability) ratios between monoisotopic peak and isotopologue peak.

```{r, eval = !file.exists(paste0(dr, "iso_data_sub_positive.RData"))}
## Calculate mass differences and intensity (probability) ratios for all
## isotopologues; putting all into a data.frame for easier data processing
iso_data <- lapply(iso_ps, function(z) {
    mono <- which(rownames(z) == "mono")
    if (length(mono) == 1L) {
        res <- data.frame(isotopologue = rownames(z),
                          md = z[, "m/z"] - z[mono, "m/z"],
                          R = z[, "abundance"] / z[mono, "abundance"],
                          z[, c("m/z", "abundance"), drop = FALSE],
                          monomass = z[mono, "m/z"],
                          check.names = FALSE)
        res[-mono, , drop = FALSE]
    } else {
        data.frame(isotopologue = rownames(z),
                   md = NA_real_,
                   R = NA_real_,
                   z[, c("m/z", "abundance"), drop = FALSE],
                   monomass = NA_real_,
                   check.names = FALSE)
    }
})
cnts <- vapply(iso_data, nrow, integer(1))
iso_data <- do.call(rbind, iso_data)
rownames(iso_data) <- NULL
iso_data$formula <- rep(adducts_frmls, cnts)
iso_data$hmdb_id <- rep(rownames(counts_df), cnts)

## subset to selected isotopologues
iso_data_sub <- iso_data[iso_data$isotopologue %in% selected_substs, ]
save(iso_data_sub, file = paste0(dr, "iso_data_sub_positive.RData"))
```

```{r, eval = file.exists(paste0(dr, "iso_data_sub_positive.RData"))}
load(paste0(dr, "iso_data_sub_positive.RData"))
```

### Finding bounds for intensity ratios on a subdivision of the x (mass) axis

Given the set of selected substitutions we want to determine, for each of them, 
mass dependent upper and lower bounds for R (ratio between the probability to 
observe the substitution in a compound and the probability to observe the 
monoisotopic substitution). In particular, we construct broken bound lines 
iteratively using the following strategy based on the available data points 
(m, R) (each of them being associated to a certain HMDB compound) for a given 
substitution. To determine an upper [lower] bound line we start from the origin
of the (mass, R) space and we select among the available points the point P1 
such that the slope of the line trough the origin and P1 is maximum [minimum]. 
Then among the points to the right of P1 (if any) we look for P2 such that the 
slope through P1 and P2 is maximum [minimum]. We proceed analogously until 
there's no point to the right of PN, i.e. PN is the point with maximum mass 
value. The broken line obtained by connecting each of the found points with 
the next one is by construction an upper [lower] bound line.
The approach seem to work relatively well for the lower bound broken line 
because the relation R vs mass tends to always be linear or superlinear. 
However, for the same reason the upper bound broken line is often the line 
through the origin and one of the rightmost points and therefore doesn't follow 
the nonlinear trend well enough. To solve the problem associated with the 
upper bound the idea is to modify the algorithm in such a way that the
next point is searched among those who have mass lower than a certain mass 
threshold rather than on all the points to the right of P.
In particular we define a subdivision `xsubd` of the mass range where points are
present. We start by searching the point P1 among those with mass s.t. 
0 < mass < `xsubd[2]`, P2 among those with mass s.t. 
P1 x value < mass < `xsubd[2]` etc.
In addition we want negative slope segments to be avoided should the algorithm 
find any. With the original algorithm negative slopes can only occur for the
the last segments of the upper bound broken line. That's because for the upper 
[lower] bound line the slope of each segment is bigger [smaller] by construction 
than that of the next one and the slope of the first segment is always 
positive (being one endpoint the origin and the other having positive 
coordinates). The modification of the original algorithm introduces the 
possibility that each segment (but the first one) can have negative slope for 
both the upper bound and lower bound line. We remove points which which cause
negative slopes segments in such a way that the remaining points still define 
a bound broken line. In case the last segments have negative slopes we replace 
the points originating them with a point at the intersection of the line 
extending the last valid (positive slope) segment and the vertical line through
the point with maximum x value.

In the following plot are reported the bounds for R vs mass obtained using the 
modified function for the upper bound. A x axis subdivision with shorter 
intervals the closer to the minimum of the mass range has been used.

```{r, echo = FALSE, fig.width = 10, fig.height = 8}
f <- rep(seq(ceiling(length(selected_substs)/9)),
         each = 9)[seq_along(selected_substs)]
plts <- split(sort(selected_substs), f)

tmp <- lapply(plts, function(z) {
  par(mfrow = c(3, 3), mar = c(4.2, 4.2, 2, 0.5))
  for (iso in z) {
    idx <- which(iso_data_sub$isotopologue == iso)
    points <- cbind(iso_data_sub$monomass[idx], iso_data_sub$R[idx])
    plot(points, main = iso, xlab = "monoisotopic mass",
         ylab = "R", pch = 16, col = "#00000040", cex = 0.75)
    xsubd <- seq(0, 1, length.out = 8)^2 * diff(range(points[, 1])) +
      min(points[, 1])
    bpl <- bound_points(points, "lower")
    points(bpl, type = "l", col = "#377EB880")
    abline(v = bpl[, 1], col = "grey", lwd = 0.5)
    bpu <- bound_points2(points, "upper", xsubd = xsubd)
    points(bpu, type = "l", col = "#4DAF4A80")
    abline(v = bpu[, 1], col = "grey", lwd = 0.5)
  }
})
```

We repeat the same plots on a portion of the mass range.
```{r, echo = FALSE, fig.width = 10, fig.height = 8}
f <- rep(seq(ceiling(length(selected_substs)/9)),
         each = 9)[seq_along(selected_substs)]
plts <- split(sort(selected_substs), f)

tmp <- lapply(plts, function(z) {
  par(mfrow = c(3, 3), mar = c(4.2, 4.2, 2, 0.5))
  for (iso in z) {
    idx <- which(iso_data_sub$isotopologue == iso)
    points <- cbind(iso_data_sub$monomass[idx], iso_data_sub$R[idx])
    #s <- which(points[, 1] <= 800)
    s <- which(points[, 1] <= min(points[, 1]) + 700)
    plot(points[s, ], main = iso, xlab = "monoisotopic mass",
         ylab = "R", pch = 16, col = "#00000040", cex = 0.75)
    xsubd <- seq(0, 1, length.out = 8)^2 * diff(range(points[, 1])) +
      min(points[, 1])
    bpl <- bound_points(points, "lower")
    points(bpl, type = "l", col = "#377EB880")
    abline(v = bpl[, 1], col = "grey", lwd = 0.5)
    bpu <- bound_points2(points, "upper", xsubd = xsubd)
    points(bpu, type = "l", col = "#4DAF4A80")
    abline(v = bpu[, 1], col = "grey", lwd = 0.5)
  }
})
```

We construct the substitution definition table.
```{r}
hmdb_subst <- lapply(selected_substs, function(sub){
  idx <- which(iso_data_sub$isotopologue == sub)
  points <- cbind(iso_data_sub$monomass[idx], iso_data_sub$R[idx])
  xsubd <- seq(0, 1, length.out=8)^2 * diff(range(points[, 1])) +
    min(points[, 1])
  bpl <- bound_points(points, "lower")
  bpu <- bound_points2(points, "upper", xsubd = xsubd)
  res <- get_bounds(bpl, bpu)
  nr <- nrow(res)
  data.frame(name = rep(sub, nr),
             md = rep(iso_data_sub$md[idx][1], nr),
             minmass = rep(min(points[, 1]), nr),
             maxmass = rep(max(points[, 1]), nr),
             res)
})
hmdb_subst <- do.call(rbind, hmdb_subst)
hmdb_subst <- hmdb_subst[order(hmdb_subst[, "md"]), ]
```

```{r}
dr_txt <- paste0("data/txt/")
dir.create(dr_txt, showWarnings = FALSE, recursive = TRUE)
write.table(hmdb_subst, 
            file = paste0(dr_txt, "hmdb_subst_positive.txt"),
            sep = "\t", row.names = TRUE)
```

The following table contains summary information about the substitutions.
```{r}
f <- factor(iso_data_sub$isotopologue)
subst_info <- data.frame(
  name = levels(f),
  minmass = vapply(split(iso_data_sub$monomass, f), min, numeric(1)),
  maxmass = vapply(split(iso_data_sub$monomass, f), max, numeric(1)),
  md = iso_data_sub$md[match(levels(f), iso_data_sub$isotopologue)],
  proportion = as.numeric(
    (subst_frequency / nrow(cmps))[match(levels(f), names(subst_frequency))]),
  proportion_var = as.numeric(
    subst_proportion_var[match(levels(f), names(subst_proportion_var))])
)
subst_info <- subst_info[order(subst_info$md), ]
rownames(subst_info) <- NULL
pandoc.table(subst_info, style = "rmarkdown")
```
# Session information

The R version and packages used in this analysis are listed below.

```{r sessioninfo}
sessionInfo()
```
